GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis - sensitivity analysis

### 2023-12-31 figs/GSE158422_sensitivity.Rmd
updates

## LRM - limma rank mean mitch

```{r,lmrm1}

dm$rank <-  rank(dm$t) - nrow(subset(dm,t<0))
mm <- merge(dm,gt,by.x=0,by.y="probe")
head(mm)
mma <-aggregate(mm$rank ~ gene,mm,mean)
rownames(mma) <- mma$gene
mma$gene = NULL
colnames(mma) <- "meanrank"
lrmres <- runmitch(m=mma,genesets=gs_symbols,cores=16)

lrmsig <- subset(lrmres,p.adjustANOVA<0.05)

top <- head(lrmsig,20)
top <- top[order(top$s.dist),]

top %>%
  kbl(caption="Top gene sets (biggest effect size with FDR<0.05)") %>%
  kable_paper("hover", full_width = F)

hist(lrmsig$s.dist)

# distinguish up (TRUE) and down (FALSE)
lrmsigsets <- paste(rownames(lrmsig),as.character(lrmsig$s.dist > 0))
lrmsigsets2 <- gsub(" TRUE"," UP",lrmsigsets)
lrmsigsets2 <- gsub(" FALSE"," DN",lrmsigsets2)

@@

top <- head(lamsig,20)
top <- top[order(top$s.dist),]

top %>%
  kbl(caption="Top gene sets (biggest effect size with FDR<0.05)") %>%
  kable_paper("hover", full_width = F)

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="LAM") ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(lamsig))

dots <- lapply(ssres,function(x) { x[,"FP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

@@

names(nsigs) <- sizes
nsigs

@@

names(ngsig) <- FCLIM
ngsig

par( mar = c(5.1, 6.1, 4.1, 2.1) )
barplot(unlist(lapply(ngsig,length)),horiz=TRUE,las=1,xlab="no. significant sets",xlim=c(0,60)) ; grid()
mtext("fold change threshold", 2, 4)

@@

Only 4 hyper and 1 hypomethylated gene set were identified.
This is surprising as there were 1299 hypermethylated probes corresponding to 808 genes, and
1016 hypomethylated probes corresponding to 657 genes.

@@

myes <- data.frame(do.call(rbind,myes))
gsiges <- data.frame(gsig,myes)
gsiges <- head(gsiges[order(-gsiges$ES),],20)
gsiges <- gsiges[order(gsiges$ES),]

@@

dots <- lapply(gsres,function(x) { x[,"TP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="GSAmeth")
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(gsig))

dots <- lapply(gsres,function(x) { x[,"FP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0)
beeswarm(dots,add=TRUE,pch=1,cex=0.6)

dots <- lapply(gsres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(gsres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="LAM") ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(lamsig))

dots <- lapply(gsres,function(x) { x[,"TP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="GSA") ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(gsig))

@@

Use Euler diagrams to find the similarity.
As medians are normally distributed about the zero, one would expect equal numbers of up and down-
methylated gene sets.
Therefore the t-test based approach could be considered better.

>>>>

@@

topup <- subset(lamres,p.adjustANOVA<0.05 & s.dist >0 )
topup <- tail(topup[order(topup$s.dist),],10)

topdn <- subset(lamres,p.adjustANOVA<0.05 & s.dist <0 )
topdn <- head(topdn[order(topdn$s.dist),],10)

topup %>%
  kbl(caption="Top gene sets hypermethylated (biggest effect size with FDR<0.05)") %>%
  kable_paper("hover", full_width = F)

topdn %>%
  kbl(caption="Top gene sets hypomethylated (biggest effect size with FDR<0.05)") %>%
  kable_paper("hover", full_width = F)

top <- rbind(topdn,topup)

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="LAM") ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(lamsig))

dots <- lapply(ssres,function(x) { x[,"FP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
lapply(ratio,summary)
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

@@

names(nsigs) <- sizes
str(nsigs)
lapply(nsigs,length)

@@

names(ngsig) <- FCLIM
str(ngsig)
lapply(ngsig,length)

par( mar = c(5.1, 6.1, 4.1, 2.1) )
barplot(unlist(lapply(ngsig,length)),horiz=TRUE,las=1,xlab="no. significant sets") ; grid()
mtext("fold change threshold", 2, 4)

@@

107 significant gene sets were identified.

@@

myes <- data.frame(do.call(rbind,myes))
gsiges <- data.frame(gsig,myes)

gsigesdn <- gsiges[grep(" DN",rownames(gsiges)),]
gsigesdn <- tail(gsigesdn[order(gsigesdn$ES),],10)
gsigesdn <- gsigesdn[nrow(gsigesdn):1,]
gsigesup <- gsiges[grep(" UP",rownames(gsiges)),]
gsigesup <- tail(gsigesup[order(gsigesup$ES),],10)

gsiges <- rbind(gsigesdn,gsigesup)

@@

dots <- lapply(gsres,function(x) { x[,"TP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",
  cex=0,main="GSAmeth",ylim=c(0,nrow(gsig)))
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(gsig))

dots <- lapply(gsres,function(x) { x[,"FP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0)
beeswarm(dots,add=TRUE,pch=1,cex=0.6)

dots <- lapply(gsres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(gsres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
lapply(ratio,summary)
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",
  cex=0,main="LAM", ylim=c(0,nrow(lamsig))) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(lamsig))

dots <- lapply(gsres,function(x) { x[,"TP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",
  cex=0,main="GSA", ylim=c(0,nrow(gsig))) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(gsig))

@@

Use Euler diagrams to find the similarity.
First, distinguish based on direction.

```{r,ol0}

lamsigsets2up <- lamsigsets2[grep(" UP", lamsigsets2)]
lamsigsets2dn <- lamsigsets2[grep(" DN", lamsigsets2)]

gsigsetsup <- gsigsets[grep(" UP", gsigsets)]
gsigsetsdn <- gsigsets[grep(" DN", gsigsets)]

v1 <- list("LAM up"=lamsigsets2up,"LAM dn"=lamsigsets2dn,
  "GSA up"=gsigsetsup, "GSA dn"=gsigsetsdn)

message("common")
intersect(lamsigsets2up, gsigsetsup)
intersect(lamsigsets2dn, gsigsetsdn)

message("specific to LAM")
LAMspec <- setdiff(lamsigsets2up, gsigsetsup)
LAMspec
LAMspec <- setdiff(lamsigsets2dn, gsigsetsdn)
LAMspec

message("specific to GSA")
GSAspec <- setdiff(gsigsetsup, lamsigsets2up)
GSAspec
GSAspec <- setdiff(gsigsetsdn, lamsigsets2dn)
GSAspec

Euler without direction noted.
